---
title: "coding advent"
output:
  html_document:
    df_print: paged
---
```{r}
library(stringr)
```


##Day 1
A. Find two numbers in the file that summed give 2020 and print their product.
```{r}
numbers <- read.delim("advent_day1.txt", stringsAsFactors = F, header=F)
numbers <- as.numeric(numbers$V1)

for (i in 1:length(numbers))
{
  difference=2020-numbers[i]
  if (length(which(numbers==difference)>0))
  {
    print(paste("a is ", numbers[i]))
    print(paste("b is ", difference))
    print(paste("product is ", numbers[i]*difference))
  }
}
```
B. Find three numbers in the file that summed give 2020 and print their product.
```{r}
for (i in 1:length(numbers))
{
  difference1=2020-numbers[i]
  for (j in 1:length(numbers)-1)
  {
    new_numbers=numbers[-i]
    difference2=difference1-new_numbers[j]
    if (length(which(new_numbers==difference2)>0))
    {
      print(paste("a is ", numbers[i]))
      print(paste("b is ", new_numbers[j]))
      print(paste("c is ", difference2))
      print(paste("product is ", numbers[i]*difference2*new_numbers[j]))
    }
  }
}
```
##Day 2
```{r}
#read in data
pass_data <- read.delim("advent_day2.txt", stringsAsFactors = F, header=F, sep=" ")
names(pass_data) <- c("interval", "char", "password")

#separate range into columns
pass_data <- tidyr::separate(pass_data, col="interval", into=c("min", "max"), sep="-", remove=T)
pass_data$min = as.numeric(pass_data$min)
pass_data$max = as.numeric(pass_data$max)

#remove : from char column
pass_data$char <- gsub(pass_data$char, pattern=":", replacement="")

#count the lines that fit the rule
correct=0
for (i in 1:nrow(pass_data))
{
  count <- str_count(pass_data$password[i], pass_data$char[i])
  #print(paste("count is ",count))
  if(count <= pass_data$max[i] && count >= pass_data$min[i])
    {
    correct=correct+1
  }
}
```

```{r}
#count the lines that fit the rule
correct=0
for (i in 1:nrow(pass_data))
{
  #break password into individual letters
  vect_pass <- unlist(strsplit(pass_data$password[i], split = ""))
  #subset the two characters occupied by positions min and max
  chars <- paste0(vect_pass[pass_data$min[i]], vect_pass[pass_data$max[i]])
  #count how many times it appears at named positions
  count <- str_count(chars, pass_data$char[i])
  #print(paste("count is ",count))
  if(count==1)
    {
    correct=correct+1
    }
}
```

##Day 3
```{r}
#make the slope wide enough to travel (about 1000 times the length of the slope)
slope <- read.delim("advent_day3.txt", stringsAsFactors = F, header=F, sep=" ")
for (i in 1:nrow(slope))
{
  slope$V1[i] <- paste(replicate(1000*nrow(slope), slope$V1[i]), collapse = "")
}

# navigate and count
vect_grab <- c()
index = 1
for(i in 2:nrow(slope))
{
  #navigate 3 paces - update index
  index=index+3
  #separate and unlist row so that indeces can be used
  vect_row <- unlist(strsplit(slope$V1[i], split = ""))
  #grab the value on the line below at the updated index
  vect_grab <- c(vect_grab, vect_row[index])
}

#print how many trees
length(which(vect_grab=="#"))
```

```{r}
# navigate and count
trees_count <- function(right, down)
{
  vect_grab <- c()
  index = 1
  #seq start 2 for down 1, 3 for down 2
  for(i in seq(3, nrow(slope), down))
    {
      #navigate 3 paces - update index
      index=index+right
      #separate and unlist row so that indeces can be used
      vect_row <- unlist(strsplit(slope$V1[i], split = ""))
      #grab the value on the line below at the updated index
      vect_grab <- c(vect_grab, vect_row[index])
    }
  print(vect_grab)
  #print how many trees
  print(length(which(vect_grab=="#")))
}

trees_count(1, 2)
trees_count(1, 1)
trees_count(7, 1)
trees_count(5, 1)
trees_count(3, 1)

66*73*85*35*272
```

##Day 4
Need to split the input file into separate files (one for each passport) based on white lines. Do with awk, make sure to close file at each iteration, and remove the DOS <CR> line terminators first.
```{bash}
tr -d '\r' < /Users/dbogdan/analysis/WORKDIR/Advent_2020/advent_day4.txt | awk 'BEGIN{i=0}{RS="";}{close("F"i);x="F"++i;}{print > x;}'
```

```{r}
#read in the docs and parse to list
file_list <- list.files(path="./sep/")
data_list <- list()
for (i in 1:length(file_list))
{
  data_list[[i]] <- read.delim(paste0("./sep/", file_list[[i]]), header=F, sep="\n")
}

#for each doc in the list, count how many fields there are and whether they have a cid field
fields <- c()
cid <- c()
for (i in 1:length(data_list))
{
  count <- 0
  cid_count <- 0
  doc <- data_list[[i]]
  for (j in 1:nrow(doc))
  {
    string <- doc$V1[j]
    count <- count + str_count(string, ":")
    cid_count <- cid_count + str_count(string, "cid")
  }
  fields <- c(fields, count)
  cid <- c(cid, cid_count)
}

summary_passports <- data.frame(n_fields=fields, country=cid)
valid <-0 
pos_valid <- c()
for (i in 1:nrow(summary_passports))
{
  #count if valid, also save position
  if (summary_passports$n_fields[i]==8)
  {
    valid <- valid+1
    pos_valid <- c(pos_valid, i)
  }
  else 
    if(summary_passports$n_fields[i]==7 && summary_passports$country[i]==0)
      {
      valid <- valid+1
      pos_valid <- c(pos_valid, i)
    }
    
}
```

```{r}
#use only valid docs from now
valid_list <- data_list[pos_valid]

#acceptable values
byr_c <- c(1920, 2002)
iyr_c <- c(2010, 2020)
eyr_c <- c(2020, 2030)
hgt_cm <- c(150, 193)
hgt_in <- c(59, 76)
ecl_p <- c("amb", "blu", "brn", "gry", "grn", "hzl", "oth")
```

```{r}
#organise data into a data frame that makes sense
byr_v <- c()
iyr_v <- c()
eyr_v <- c()
hgt_v <- c()
ecl_v <- c()
hcl_v <- c()
pid_v <-c()
for (i in 1:length(valid_list))
{
  doc <- valid_list[[i]]
  doc_string <- c("")
  for (j in 1:nrow(doc))
  {
    doc_string <- paste(doc_string, doc$V1[j])
  }
  #split into different strings for each attribute, then for each value
  doc_string <- str_split(doc_string, " ")
  att <- c()
  for (n in 1:length(doc_string[[1]]))
  {
    att <- c(att, str_split(doc_string[[1]][n], ":")[[1]][1], 
             str_split(doc_string[[1]][n], ":")[[1]][2])
    
  }
  #find the 7 attributes and grab the values 
  byr_v <- c(byr_v, att[[which(att=="byr")+1]])
  iyr_v <- c(iyr_v, att[[which(att=="iyr")+1]])
  eyr_v <- c(eyr_v, att[[which(att=="eyr")+1]])
  hgt_v <- c(hgt_v, att[[which(att=="hgt")+1]])
  ecl_v <- c(ecl_v, att[[which(att=="ecl")+1]])
  hcl_v <- c(hcl_v, att[[which(att=="hcl")+1]])
  pid_v <-c(pid_v, att[[which(att=="pid")+1]])
}

#make data frame
all_entries <- data.frame(byr=byr_v, iyr=iyr_v, eyr=eyr_v, hgt=hgt_v, ecl=ecl_v, hcl=hcl_v, pid=pid_v)

checked <- all_entries %>% dplyr::filter(byr >= byr_c[1] & byr <= byr_c[2]) %>% dplyr::filter(iyr >= iyr_c[1] & iyr <= iyr_c[2]) %>% dplyr::filter(eyr >= eyr_c[1] & eyr <= eyr_c[2])

checked <- checked[grep("#", checked$hcl),]
checked$hcl <- gsub("#", "", checked$hcl)
#check if there are any that do not fit the 6 digit rule
which(str_length(checked$hcl)!=6)

#exclude any non-9 digit pid numbers
checked <- checked[-which(str_length(checked$pid)!=9),]

#deal with eye colour
indeces <- c()
for (i in 1:length(checked$ecl))
{
  if(length(setdiff(ecl_p, checked$ecl[i]))<7)
  {
    indeces <- c(indeces, i)
  }
}

checked <-checked[indeces,]

#deal with height
numbers <- gsub("[^[:digit:]]","",checked$hgt)
unit <- gsub("[[:digit:]]","", checked$hgt)
fin <- 0

#loop and count matches
for (i in 1:length(numbers))
{
  if (unit[i]=="cm")
  {
    if (numbers[i]>=hgt_cm[1] & numbers[i]<=hgt_cm[2])
    {
      fin=fin+1
    }
  }
  else 
  {
    if (numbers[i]>=hgt_in[1] & numbers[i]<=hgt_in[2])
    {
      fin=fin+1
    }
  }
}
```

##Day 5.
```{r}
seats <- read.delim("advent_day5.txt", header=F, sep="\n")
rs=128
cs=8

all_ids <- c()
row_ids <- c()
col_ids <- c()
for (seat in  1:nrow(seats))
{
  s=seats$V1[seat]
  all_ids <- c(all_ids, seat_id(s)[3])
  row_ids <- c(row_ids, seat_id(s)[1])
  col_ids <- c(col_ids, seat_id(s)[2])
}


#write function that calculates seat ids
seat_id <- function(s)
{
  #split string to be able to access individual positions
  s=str_split(s, "")[[1]]
  #initialise up and down boundaries of interval to be halved
  dn=0
  up=rs
  #initialise the row position to be updated
  fin_r=rs/2
  i=1
  #deal with the row
  while (length(s)>3)
  {
    if (s[i]=="F")
      {
        up <- fin_r
        dn <- dn
        fin_r <- dn + floor((up-dn)/2)
      }
    else 
      {
        up <- up
        dn <- fin_r
        fin_r <- dn + floor((up-dn)/2) 
      }
    #cut the letter from vect
    s=s[2:length(s)]
  }
  
  #deal with the column -same procedure
  dn=0
  up=cs
  fin_c=cs/2
  for (i in 1:3)
  {
    if (s[i]=="L")
      {
        up <- fin_c
        dn <- dn
        fin_c <- dn + floor((up-dn)/2)
      } 
    else 
      {
        up <- up
        dn <- fin_c
        fin_c <- dn + floor((up-dn)/2) 
      }
  }
  #calculate seat id
  id <- fin_r*8 + fin_c
  #return vector with ids found
  return(c(fin_r, fin_c, id))
}
```

```{r}
#put rows and columns in data frame
x = data.frame(row=row_ids, col=col_ids)
x = x[order(x$row),]
#collapse into counts for each row
x1 <- x %>% dplyr::group_by(row) %>% dplyr::summarise(seats = paste(row, collapse=", "), times = length(row)) 
x1 = x1[order(x1$row),]
#find my seat
which(x1$times<8)
```

##Day 6
Collapse master file so that each group has its own line.
```{bash}
awk -v RS= '$1=$1' advent_day6.txt > day6_collapsed.txt
```

Count questions ANYONE answered yes (unique letters).
```{r}
data_list <- read.delim("/Users/dbogdan/analysis/WORKDIR/Advent_2020/day6_collapsed.txt", sep="\n", header=F)

all_counts <- c()
for (i in 1:nrow(data_list))
{
  count <- sum(!!str_count(data_list$V1[i], letters))
  all_counts <- c(all_counts, count)
}

sum(all_counts)
```

Count questions EVERYONE answered yes (intersect strings within each line).
```{r}
all_counts <- c()
for (i in 1:nrow(data_list))
{
  row <- str_split(data_list$V1[i], " ")[[1]]
  count <- length(Reduce(intersect, strsplit(row,"")))
  all_counts <- c(all_counts, count)
}

sum(all_counts)
```

##Day 7
Tidy up the data file in sed before importing to R
```{bash eval=FALSE}
cat advent_day7.txt | sed 's/ contain/,/g' | sed 's/ bags//g' | sed 's/\.//g' > fin.txt
```

```{bash eval=FALSE}
#increase R memory - not a permanent change
ulimit -s 16384 # enlarge stack limit to 16 megs

#check R memory
R --slave -e 'Cstack_info()["size"]'
```

This code needs to be run in a separate R session, after the memory requirement has been expanded from the command line as above.
```{r eval=FALSE}
#read individual lines from data
bags_data <- read.delim("/Users/dbogdan/analysis/WORKDIR/Advent_2020/fin.txt", header = FALSE, sep = "\n")
split_data <- c()
for (i in 1:nrow(bags_data))
{
  row <- str_split(bags_data$V1[i],"\\,",n=2)
  split_data <- rbind(split_data, row[[1]])
}

row.names(split_data) <- split_data[,1]
split_data <- split_data[,2]


f_recursion <- function(colour, split_data)
{
  subset <- split_data[grep(split_data, pattern = colour)]
  if (length(subset)>0)
  {
      for (i in 1:length(subset))
        {
        colour <- names(subset[i])
        print(colour)
        f_recursion(colour, split_data)
        }
  }
}

colour <- "shiny gold"
f_recursion(colour, split_data)
```

The recursive function will print a list of colours to the console. Save this list to a file then load it in here to find the unique colours:
```{r}
unique_cols <- read.delim("/Users/dbogdan/analysis/WORKDIR/Advent_2020/7_part1_interim.txt", sep=" ", header=F)
length(unique(unique_cols$V2))
```

Part 2.
```{r}
f_recursion_2 <- function(colour, n)
{
  subset <- split_data[colour]
  if (subset[1]==" no other")
      print(paste(colour, n))
  else
  {
      split_string <- str_split(subset[1], ",")[[1]]
      numbers <- as.numeric(gsub(".*?([0-9]+).*", "\\1", split_string))
      sum <- n*sum(numbers) + n
      print(numbers)
      print(paste(colour, sum))
      if (length(split_string)>0)
      {
        for (i in 1:length(split_string))
        {
          colour <- substring(str_split(split_string, pattern=",")[[i]], 4)
          n <- numbers[i]
          f_recursion_2(colour, n)
        }
      }
  }
}


f_recursion_2(colour, 1)
```

```{r}
unique_cols <- read.delim("/Users/dbogdan/analysis/WORKDIR/Advent_2020/7_part2_interim.txt", sep=" ", header=F)
bags_to_sum <- unique(unique_cols$V2)
counts <- as.numeric(gsub(".*?([0-9]+).*", "\\1", bags_to_sum))
sum(counts)
```

##Day 8
VM problem -come back to this and make it nice because likely to come back and bite you in the ass
```{r}
loop <- read.delim("/Users/dbogdan/analysis/WORKDIR/Advent_2020/advent_day8.txt", header = FALSE, sep = " ")
loop_vect <- loop$V2
loop_vect <- as.numeric(loop_vect)
names(loop_vect) <- loop$V1

next_pos=1
this_pos=1
a=0
while (loop_vect[next_pos] < 1000004)
{
  this_pos <- next_pos
  if (names(loop_vect)[this_pos]=="nop")
    {
    loop_vect[this_pos] <- 1000000
    next_pos=this_pos+1
    }
  else
  {
    if (names(loop_vect)[this_pos]=="acc")
    {
      a=a+loop_vect[this_pos]
      loop_vect[this_pos] <- 1000000
      next_pos=this_pos+1
    }
    else
    {
      x=loop_vect[this_pos]
      loop_vect[this_pos] <- 1000000
      next_pos=as.numeric(x)+this_pos
    }
  }
}

##Answer 1
a
```

Part 2.
```{r}
find_a <- function(loop_vect)
{
  next_pos=1
  this_pos=1
  a=0
  repeat
  {
    this_pos <- next_pos
    if (names(loop_vect)[this_pos]=="nop")
      {
      loop_vect[this_pos] <- 1000000
      next_pos=this_pos+1
      }
    else
    {
      if (names(loop_vect)[this_pos]=="acc")
      {
        a=a+loop_vect[this_pos]
        loop_vect[this_pos] <- 1000000
        next_pos=this_pos+1
      }
    else
    {
      x=loop_vect[this_pos]
      loop_vect[this_pos] <- 1000000
      next_pos=as.numeric(x)+this_pos
    }
  }
  if (next_pos > length(loop_vect)+1)
    break;
}

return(a)
}

#prepare data in named vector
loop_vect <- loop$V2
loop_vect <- as.numeric(loop_vect)
names(loop_vect) <- loop$V1

##this loop will get stuck on the right substitution because names(loop_vect)[this pos]
##in the find_a function will be undefined

##running this code will result in the right new_loop_vect that terminated normally
for (i in 1:length(loop_vect))
{
  if (names(loop_vect)[i]=="nop")
  {
    new_loop_vect <- loop_vect
    names(new_loop_vect)[i] <- "jmp"
    a = find_a(new_loop_vect)
  }
  else
  {
    if (names(loop_vect)[i]=="jmp")
    {
    new_loop_vect <- loop_vect
    names(new_loop_vect)[i] <- "nop"
    a = find_a(new_loop_vect)
    }
  }
}

```

```{r}
#finally, run the finding a algorithm on the new loop vect, substituted at the i position
  loop_vect <- new_loop_vect
  next_pos=1
  this_pos=1
  a=0
  repeat
  {
    this_pos <- next_pos
    if (names(loop_vect)[this_pos]=="nop")
      {
      loop_vect[this_pos] <- 1000000
      next_pos=this_pos+1
      }
    else
    {
      if (names(loop_vect)[this_pos]=="acc")
      {
        a=a+loop_vect[this_pos]
        loop_vect[this_pos] <- 1000000
        next_pos=this_pos+1
      }
    else
    {
      x=loop_vect[this_pos]
      loop_vect[this_pos] <- 1000000
      next_pos=as.numeric(x)+this_pos
    }
  }
  if (next_pos > length(loop_vect)+1)
    break;
  }
#Answer 2
a
```

##Day 9
```{r}
code <- read.delim("/Users/dbogdan/analysis/WORKDIR/Advent_2020/advent_day9.txt", header = FALSE, sep = "\n")
code <- as.numeric(code$V1)

i=26
pass=TRUE
while (pass==TRUE)
{
  previous <- code[(i-25):(i-1)]
  for (j in 1:length(previous))
  {
    value = abs(previous[j]-code[i])
    if(length(which(previous==value))>0)
    {
      pass=TRUE
      break;
    }
    else
    {
      pass=FALSE
    }
  }
  i=i+1
}

##Answer 1
code[i-1]
```

Part 2
```{r}
target=code[i-1]
pass=TRUE
i=1
while (pass==TRUE)
{
  sum=0
  for (j in i:length(code))
  {
    sum=sum+code[j]
    if (sum==target)
    {
      pass=FALSE
      break;
    }
    else
    {
      if (sum > target)
      {
        pass=TRUE
        break;
      }
    }
  }
  i=i+1
}

#check the sum is right
sum(code[(i-1):j])==target

##Answer 2
min(code[(i-1):j])+max(code[(i-2):j])
```

